---
title: "Introduction to BoeckLabRtools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to BoeckLabRtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The BoeckLabRtools package provides utilities for analyzing data from experiments in the Boeck Lab. This vignette demonstrates a common workflow: analyzing ASCT (Automated Single-Cell Tracking) killcurves to determine antimicrobial efficacy.

ASCT killcurves track cell death over time by monitoring the uptake of propidium iodide (PI), a fluorescent dye that penetrates dead cells. The resulting data allows us to calculate the area under the curve (AUC) as a measure of antimicrobial effectiveness.

# Example analysis of ASCT killcurves generated with our pipelines
The goal of this exercise is to show how to generate kill curves and calculate the AUC from the raw output of the ASCT python pipeline. For this exercise, we will use the example data ported with the package. This data is a subset of the data generated by Mrs Sara Topra, a former master student in our lab. The first step is to load the basic libraries and the data itself

```{r load_libraries}
library(BoeckLabRtools)
library(data.table)
library(ggplot2)
library(magrittr)
d <- data.table::as.data.table(BoeckLabRtools::killcurves)
```

The data is a data.table with the following columns: `colnames(d)`
- **file**: Character vector representing the file name
- **frame**: Integer vector representing the frame number
- **channel**: Integer vector representing the channel number
- **date_time**: POSIXct vector with the date and time of the frame
- **timestep**: Numeric vector representing the time step from the start of the experiment
- **object_class**: Character vector representing the object class (single-cell, noise, off-focus, or clump)
- **total_count**: Integer vector with the total count of objects for the given class
- **pi_class_neg**: Integer vector with the PI negative class count
- **pi_class_pos**: Integer vector the PI positive class count
- **area_pineg**: Numeric vector of the total area of PI negative objects
- **area_pipos**: Integer vector of the total area of PI positive objects

Next, we will try to extract the well and field information that is stored in the filenaming pattern. For that, we use the `tstrsplit` function from the data.table package. The `tstrsplit` function is a powerful function that allows us to split strings based on a regular expression pattern. In this case, we will split the file name based on the underscore character (`_`). The `keep` argument is used to keep the second and fourth elements of the split. The result is stored in the `well` and `field` columns.

```{r extract_well_info}
d[, c("well", "field") := tstrsplit(file, "_", keep = c(2, 4))]
d[, c("row", "col") := data.table::tstrsplit(well, "(?<=\\D)(?=\\d)", perl = TRUE)]
d[, col := as.integer(col)]
```

## Filling Missing Classes

The next step is to fill the missing classes. This is due to the fact that the segmentation and classification algorithm sometimes does not find object of some class in a particular frame. This is normal, but not having all the objects for every frame can cause problems downstream. The `fill_missing_classes` function is a helper function that fills the counts of missing classes with 0 values. The `key_cols` argument is a character vector with the columns that will be used to group the data. The `fill_col` argument is the column that will be filled. The `cols_0count` argument is a character vector with the columns that will be filled with zeros. The `required_values` argument is a character vector with all the unique values that will be used to fill the missing classes.

```{r fill_time_cols}

d <- fill_missing_classes(
  dt = d,
  key_cols = c("file", "frame", "row", "col", "well", "field"),
  fill_col = "object_class",
  cols_0count = c("total_count", "pi_class_neg", "pi_class_pos", "area_pineg", "area_pipos"),
  required_values = unique(d$object_class)
)
```
Now, given the fact that we have inputed some rows, we also need to fill the data_time and timestep (and maybe other columns), but instead of filling with 0, we fill with the unique time for that frame and file. We take the 2 value because the first value is the NA itself.

```{r}
fill_cols_uniq_val <- c("date_time", "timestep")
d[, (fill_cols_uniq_val) := lapply(.SD, function(x) fifelse(is.na(x), unique(x)[2], x)),
  .SDcols = fill_cols_uniq_val, 
  by = .(file, frame)]
```

## Adding Treatment Information

The data is now ready to include the drug info or species info from our plate schema. If your plate schema looks as follows:
```{r table2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tbl <- "| row | 3   | 4   | 5   | 6   | 7   | 8   | 9   | 10  | 11  | 12  |
|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|-----|
| C   | IPM | TGC | TGC | AMK | AMK | AMK | TGC | TGC | MXF | MXF |
| D   | IPM | TGC | TGC | AMK | AMK | AMK | TGC | IPM | CLR | CLR |
| E   | IPM | TGC | TGC | AMK | TGC | TGC | TGC | IPM | CLR | TGC |
| F   | IPM | AMK | AMK | TGC | TGC | TGC | TGC | IPM | TGC | AMK |
| G   | MXF | AMK | TGC | TGC | TGC | TGC | TGC | TGC | AMK | TGC |
| H   | TGC | AMK | AMK | TGC | TGC | TGC | TGC | IPM | AMK | AMK |
"
cat(tbl)
```
Then, you can use the `add_well_info` function to include the drug info in the data table. In this example, we use the already example plate schema, but you can add the data directly from the file
```{r add_treatment_info}

plate_schema <- data.table(BoeckLabRtools::plate_schema) # This is the example schema for our dataset
d <- add_well_info(d, 
                   plate_schema, # In your case plate schema can be a path to the file
                   info_name = "treatment",
                   plate_format = "plate_view"
                   )
```

## Calculating Live Cell Fraction

At this point, we can proceed to calculate the relative live fraction. For that, we will focus on the single-cell objects and then divide the counts of PI cells by the total number of single-cells (PI + no PI single cell). This, give us the fraction of PI positive cells (dead cells). The live fraction is then calculated as 1 - PI fraction.

```{r calculate_live_fraction}

d[, valid_cell := data.table::fifelse(object_class %in% c('single-cell'), TRUE, FALSE)]
d_sc <- d[object_class == "single-cell"]
d_sc[, pi_frac := pi_class_pos/total_count, 
        by = .(file, valid_cell, frame, object_class)]
d_sc[, live_frac := 1 - pi_frac]

```

## Interpolating Time Points

Finally, we need to get the same time point for all our frames. For large experiments, it is very common to have a shift ranging from minutes to hours from the wells that are scanned early in the loop to the later wells. To correct this, we use the `interpolate_variable` function. This function interpolates the variable of interest (in this case, the live fraction) for the time points of interest calculating the linear mid-point between two given points. The `timepoints` argument is the time points that we want to interpolate. We can set plain regular intervals, or a vector of timepoints of interest (0, 6, 12, 24...). The `fill_extremes` argument is used to fill the first and last time points with the first and last values of the variable of interest. The `return_time` argument is used to return the time points of interest, if we want to interpolate more than one variable, we can set this to `FALSE` for subsequent calls. The result is stored in the `interpol_time` and `interpol_live_frac` columns. Note the use of `by =` in order to denote that we interpolate each live fraction by each file. After that, it is also important to consider that we use the `enforce_decrease` in order to make sure that the live fraction is always decreasing.

```{r interpolate_time_points}

d_sc[, c("interpol_time", "interpol_live_frac") := interpolate_variable(time = timestep, 
                                                                     values = live_frac, 
                                                                     timepoints = c(0, 3, 6, 9, 12), 
                                                                     fill_extremes = TRUE, 
                                                                     return_time = TRUE),
          by = .(file)]

d_sc[, interpol_live_frac := enforce_decrease(interpol_live_frac), 
  by = .(file)]

```
At this point, we have our kill curves! We only need to plot them. Depending on the size of your dataset, you can do a simple grouped plot by panel or if you have too many conditions, you might need to filter your data.

## Visualizing the Kill Curves


```{r plot_killcurves, fig.width=10, fig.height=8}
# plot
d_sc %>%
  ggplot(., aes(x = timestep, y =live_frac, group = file)) +
  geom_point() +
  facet_wrap(well ~ .) +
  theme_BoeckLab()

```

## Calculating Area Under the Curve (AUC)

For further analysis, we can compress the data to a single value. For that, we can calculate the area under the curve (AUC) for each well. The `calculate_AUC` function calculates the AUC for each well. The `key_cols` argument is a character vector with the columns that will be used to group the data. The `value_col` argument is the column that will be used to calculate the AUC. The `time_col` argument is the column that will be used as the x-axis. The result is stored in the `AUC` column. In this case, we group by as many variables as we need using the `by =` argument.

```{r calculate_auc}

auc_data <- d_sc[, .(auc = calc_auc(x = interpol_time, y = interpol_live_frac)), 
              by = .(file, well, field, row, col, treatment)]

```

## Summarizing Results by Treatment

As last example figure, we can make a bar plot with error bars for our AUC for each treatment:
```{r summarize_by_treatment, fig.width=8, fig.height=6}

auc_data %>%
  .[, .(auc = mean(auc), sd_auc = sd(auc)), 
    by = .(treatment)] %>%
  ggplot(., aes(x = treatment, y = auc, fill = treatment)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(ymin = auc - sd_auc, ymax = auc + sd_auc), 
                position = position_dodge(width = 0.9), width = 0.25) +
  theme_BoeckLab()
```



