---
title: "QC Report"
output: 
  pdf_document:
    fig_caption: yes
    fig_height: 8
    fig_width: 10
params:
  file: ""
---


```{r setup, include=FALSE, echo=FALSE}
# Load required libraries
library(data.table)
library(ggplot2)
library(gridExtra)
library(viridis)

# Error if file not found
if (!file.exists(params$file)) {
  stop("File not found.")
}

# Import data
d <- fread(params$file)
d[, c("experiment", "well", "field") := tstrsplit(file, "_", keep = c(1, 3, 4))]
d[,  well := rename_well(well, file)]
d[, c("row", "col") := tstrsplit(well, "(?<=\\D)(?=\\d)", perl=TRUE)]
d[, col := as.integer(col)]
```
# Object Class Distribution

This dataset contains `r length(unique(d$file))` unique fields and `r length(unique(d$well))` wells out of a total possible 1536 wells. The fraction of used wells is `r sprintf("%.2f%%", length(unique(d$well)) / 1536 * 100)`.

This plot illustrates the spatial distribution of object classes across the plate at the initial time point (t0). The visualization aggregates data by summing the counts for each object class across all fields within each well. Typically, each well comprises 9 fields, providing a comprehensive view of object distribution.

The unique object classes present in this dataset are: `r paste(unique(d$object_class), collapse = ", ")`.

For each object class, a heatmap is generated where:
- The x-axis represents the plate columns
- The y-axis represents the plate rows
- The color intensity indicates the total count of objects in each well

This visualization allows for quick identification of patterns or anomalies in object distribution across the plate, potentially highlighting experimental inconsistencies or biological phenomena of interest.

```{r, echo = FALSE}
  object_class_plots <- create_object_class_plots(d)
```

\newpage
# Growth Plot
The growth plot provides a comprehensive overview of the growth patterns observed across all wells throughout the experiment. 

Key aspects of this plot:

1. Growth Ratio Calculation: For each well, we compute the ratio of the maximum total area to the minimum total area across all time points. This ratio serves as a proxy for growth.

2. Logarithmic Scale: The growth ratio is log-transformed to better visualize the wide range of growth patterns, from minimal to extensive growth.

```{r, echo = FALSE}
growth_plot <- create_growth_plot(d)
plot(growth_plot)
```

\newpage

# Consistency Plot
This plot demonstrates the consistency of classification in the birght field across different conditions.
Key features of this plot:

1. Percent Change Calculation: For each well and object class, a linear regression model is fitted to the object count over time. The slope of this model, normalized by the initial count and expressed as a percentage, represents the rate of change in classification.

2. Multiple Object Classes: The plot is faceted by object class, allowing comparison of classification consistency across different types of objects.


```{r, echo=FALSE}
growth_plot <- create_consistency_plot(d)
plot(growth_plot)
```
\newpage

# PI Consistency Plot
This plot visualizes the consistency of Propidium Iodide (PI) classification across different wells and time points, focusing on single-cell objects.

Key aspects of this plot:

1. PI Fraction Calculation: For each well, frame, and object class, the fraction of PI-positive cells is calculated.

2. Live Fraction: The complement of the PI fraction represents the live cell fraction.

3. Standard Deviation: For each well, the standard deviation of the live fraction across all time points is computed. This serves as a measure of classification consistency over time.

```{r, echo=FALSE}
pi_consistency_plot <- create_pi_consistency_plot(d)
plot(pi_consistency_plot)
```
\newpage

# Time Delay Plot
This plot illustrates the time delays observed for the imaging in the first frame across the plate.

```{r, echo=FALSE}
time_delay <- create_time_delay_plot(d)
plot(time_delay)
```